The Effect of a Second AIDS Defining Illness on Survival in the MACS

### 2015-05-05 paper.Rmd
First stab at prop scores

We read in the data.
```{r data}
# read in horiz data with lots of covariates
covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
           # date of death
    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
           # date of aids diagnosis
           aids.date=as.Date("1960-01-01") + FRSTAIDD,
           # date of secondary aids diagnosis
           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
           # what was the first aids diagnosis
           frst.aids.diag=AIDSDIAG,
           # what was the second aids diagnosis
           sec.aids.diag=SECAIDSDIAG,
           # did the person have a sec aids diag
           had.sec=year(sec.aids.date) < 2100,
           # did the person die
           died=year(death.date) < 2100,
           age=as.Date("1960-01-01") + DOB,
           exit=ifelse(!died, as.Date("1994-12-31"), death.date),
           censor.date=as.Date("1970-01-01") + as.integer(exit),
           last.alive=as.Date("1960-01-01") + LASTALID,
           censor.date=ifelse(last.alive < censor.date, last.alive, 
                              censor.date),
           censor.date=as.Date("1970-01-01")+censor.date,
           dead.at.censor=(death.date < as.Date("1994-12-31"))) %>%
    select(STUDYID, 
           aids.date,
           sec.aids.date,
           frst.aids.diag, 
           censor.date, 
           sec.aids.diag,
           dead.at.censor, 
           death.date) %>%
    filter(year(aids.date) < 2100)

# read in visit level data
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT==1) %>%
    select(STUDYID) %>%
    distinct(STUDYID)

@@

```{r fig1}
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # put visits in correct 1,2, 3,... form
    mutate(visit=VISIT / 10,
           visit.date=mdy(VDATE),
           haart.era.visit=year(visit.date) <= 1994) %>%

@@

# what about therapy?
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT == 1) %>% 
    group_by(STUDYID) %>%
    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
                           POPUSE, na.rm=TRUE) > 1,
              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
                                                         16, 23)) %>%
    ungroup() %>%
    # remove the 8 missing depression status
    filter(!is.na(depression))

# cd4 closest to primary AIDS diagnosis
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    filter(COHORT == 1, !is.na(CD4N)) %>%
    select(STUDYID, cd4=CD4N, VDATE) %>%
    mutate(VDATE=mdy(VDATE),
           VDATE=as.Date(VDATE)) %>%
    inner_join(covariates, by=c("STUDYID"="STUDYID")) %>%
    select(STUDYID, cd4, VDATE, aids.date) %>%
    mutate(visit.aids.diff=difftime(VDATE, aids.date)) %>%
    group_by(STUDYID) %>%
    filter(abs(visit.aids.diff) == min(abs(visit.aids.diff))) %>%
    select(STUDYID, cd4) %>%
    ungroup() %>%
    right_join(visits)

@@

```{r propensity-score, eval=FALSE}
prop.score <- glm(sec.aids ~ factor(race) + age.at.aids + drug.use +
                  aids.year + depression + cd4,
                  family=binomial,
                  data=time.varying)

time.varying <- time.varying %>%
    mutate(propensity=predict(prop.score, type="response"))
```

```{r matching, eval=FALSE}
library(MatchIt)
```

```{r sophisticated-hazards}
```

```{r graphs}
```

>>>>

We read in the data.
```{r data}
# read in horiz data with lots of covariates
covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv", 
                       progress=FALSE) %>%
           # date of death
    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
           # date of aids diagnosis
           aids.date=as.Date("1960-01-01") + FRSTAIDD,
           # date of secondary aids diagnosis
           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
           # what was the first aids diagnosis
           frst.aids.diag=AIDSDIAG,
           # what was the second aids diagnosis
           sec.aids.diag=SECAIDSDIAG,
           # did the person have a sec aids diag
           had.sec=year(sec.aids.date) < 2100,
           # did the person die
           died=year(death.date) < 2100,
           age=as.Date("1960-01-01") + DOB,
           exit=ifelse(!died, as.Date("1994-12-31"), death.date),
           censor.date=as.Date("1970-01-01") + as.integer(exit),
           last.alive=as.Date("1960-01-01") + LASTALID,
           censor.date=ifelse(last.alive < censor.date, last.alive, 
                              censor.date),
           censor.date=as.Date("1970-01-01")+censor.date,
           dead.at.censor=(death.date < as.Date("1994-12-31"))) %>%
    select(STUDYID, 
           aids.date,
           sec.aids.date,
           frst.aids.diag, 
           censor.date, 
           sec.aids.diag,
           dead.at.censor, 
           death.date) %>%
    filter(year(aids.date) < 2100)

# read in visit level data
visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT==1) %>%
    select(STUDYID) %>%
    distinct(STUDYID)

@@

```{r fig1}
visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # put visits in correct 1,2, 3,... form
    mutate(visit=VISIT / 10,
           visit.date=mdy(VDATE),
           haart.era.visit=year(visit.date) <= 1994) %>%

@@

# what about therapy?
visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT == 1) %>% 
    group_by(STUDYID) %>%
    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
                           POPUSE, na.rm=TRUE) > 1,
              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
                                                         16, 23)) %>%
    ungroup() %>%
    # remove the 8 missing depression status
    filter(!is.na(depression))

# cd4 closest to primary AIDS diagnosis
visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    filter(COHORT == 1, !is.na(CD4N)) %>%
    select(STUDYID, cd4=CD4N, VDATE) %>%
    mutate(VDATE=mdy(VDATE),
           VDATE=as.Date(VDATE)) %>%
    inner_join(covariates, by=c("STUDYID"="STUDYID")) %>%
    select(STUDYID, cd4, VDATE, aids.date) %>%
    mutate(visit.aids.diff=difftime(VDATE, aids.date)) %>%
    group_by(STUDYID) %>%
    filter(abs(visit.aids.diff) == min(abs(visit.aids.diff))) %>%
    select(STUDYID, cd4) %>%
    ungroup() %>%
    right_join(visits)

@@

Finally, we use propensity score matching to account for confounding bias.

```{r propensity-score}
set.seed(90210)
match.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
                              age.at.aids + drug.use +
                              aids.year + depression + cd4,
                              data=time.varying,
                              distance="logit",
                              method="nearest",
                              caliper=.07,
                              discard="both")

# ## let's check out the diagnostics

# ## numerical check
match.check <- summary(match.set, standardize=TRUE)

# ## graphical check
plot(match.set, type="hist")
# plot(match.set, type="jitter", interactive=FALSE)

matched.dat <- MatchIt::match.data(match.set,"all")

ph.match <- coxph(Surv(start, stop, status) ~ sec.aids +
                  factor(race) + age.at.aids + drug.use +
                  aids.year + depression + cd4,
   				  data=matched.dat,
                  method="efron")
```

There were `r length(unique(matched.dat$STUDYID))` individuals in the 
matched data set of the orginal `r nrow(data)`. There are 
`r sum(matched.dat$sec.aids)` events in the matched data set.
We perform a check on the data and see the following summaries.

```{r}
knitr::kable(match.check$sum.all,
             caption="Summary of balance for all data")
knitr::kable(match.check$sum.matched,
             caption="Summary of balance for matched data")
knitr::kable(match.check$reduction,
             caption="Percent Balance Improvement")
knitr::kable(match.check$nn,
             caption="Sample Sizes")
```

In recalculating our proportional hazards model using propensity score
in order to handle confounding bias, we see the following results.

```{r}
knitr::kable(coef(summary(ph.match)),
             caption="PH Model (Propensity Scores)")
```

```{r}
# ########################################################################
# ## Q10b Get HRs after matching with higher caliper
# ########################################################################
# 
# 
# set.seed(1428)
# match2.set <- matchit(hivhrt~age + bmi,data=dat,distance=dat$cpoe,method="nearest",caliper=.5,discard="both")
# 
# ## let's check out the diagnostics
# 
# ## numerical check
# print(sum.match2 <- summary(match2.set,standardize=T))
# 
# ## graphical check
# plot(match2.set,type="jitter",interactive=F)
# plot(sum.match2,interactive=F)
# 
# matched2.dat <- match.data(match2.set,"all")
# 
# ## how many individuals in matched.dat?
# nrow(matched2.dat)
# 
# ##
# sum(matched2.dat$dm)
# 
# ph.match2 <- coxph(Surv(time=t,event=dm)~hivhrt,data=matched2.dat,method="efron")
# summary(ph.match2)
# 
# 
# 
# 
# ########################################################################
# ## Q10c Get HRs after subclassification
# ########################################################################
# 
# 
# dat$hivhrtn <- as.numeric(dat$hivhrt)-1
# sub.set <- matchit(hivhrtn~age + bmi,data=dat,method="subclass",sub.by="all",subclass=5,discard="both")
# 
# summary(sub.set,standardize=T)
# plot(sub.set,type="jitter",interactive=F)
# plot(summary(sub.set,standardize=T),interactive=F)
# 
# sub.dat <- match.data(sub.set,"all")
# ph.sub <- coxph(Surv(time=t,event=dm)~hivhrtn+strata(subclass),data=sub.dat,method="efron")
# summary(ph.sub)
```

```{r graphs}
plot(survfit(ph.match))
```
